Parameter Recovery Validation

### 2025-06-15 nbs/concepts/parameter-recovery-validation/index.qmd
docs(parameter-recovery-validation): replace `t^*_{0,\text{on}i}` with `t^*_{on,i}`
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

**Activation Onset Times**: $t^*_{0,\text{on}i} \sim \text{Normal}(\mu_{on}, \sigma_{on}^2)$

- **Dimensionless interpretation**: Gene $i$ activates at $t^*_{0,\text{on}i}$ characteristic timescales after process initiation
- **Real-time interpretation**: Gene $i$ activates at $t^*_{0,\text{on}i}/\alpha_{\text{ref}}$ hours after process start
- **Relative interpretation**: Gene $i$ activates at $(t^*_{0,\text{on}i}/T_M) \times 100\%$ through the process

**Example interpretations:**
- $t^*_{0,\text{on}i} = 1.5$, $T_M = 5.0$: Gene activates 30% through the process
- $t^*_{0,\text{on}i} = -0.5$: Gene was already active 0.5 timescales before observation began

@@

- Posterior: $T_M \approx 6.0 \pm 0.5$
- Inferred characteristic timescale: $48/6 = 8$ hours
- Gene with $t^*_{0,\text{on}i} = 2.0$: Activates at $2.0 \times 8 = 16$ hours
- Gene with $\delta^*_i = 1.5$: Active for $1.5 \times 8 = 12$ hours

@@

- $t^*_{0,\text{on}i}$: When a gene's regulatory program activates (intrinsic timing)
- $\delta^*_i$: How long a gene's regulatory program lasts (intrinsic duration)

@@

\begin{align}
t^*_{0,\text{on}i} &\sim \text{Normal}(1.5, 0.8^2) \quad \text{(fixed biological timing)} \\
\delta^*_i &\sim \text{LogNormal}(0.0, 0.45^2) \quad \text{(fixed biological duration)}
\end{align}

**NOT dependent on $T_M$:**
\begin{align}
t^*_{0,\text{on}i} &\not\sim \text{Normal}(0.3 \times T_M, 0.8^2) \quad \text{(would create artificial scaling)}
\end{align}

@@

1. **Gene parameters are sampled independently**: $t^*_{0,\text{on}i} \not\propto T_M$ and $\delta^*_i \not\propto T_M$

2. **Scaling cell times but not gene parameters changes piecewise phase assignments**:
   - Original: Cell $j$ in phase determined by comparing $t^*_j$ vs. $t^*_{0,\text{on}i}$
   - Scaled: Cell $j$ in phase determined by comparing $\lambda t^*_j$ vs. $t^*_{0,\text{on}i}$ (unchanged)

3. **The piecewise analytical solutions depend on comparisons like $t^*_j$ vs. $t^*_{0,\text{on}i}$**:
   - Phase boundaries change when cell times are scaled but gene times are not
   - Different phase assignments lead to different analytical solution forms

4. **Therefore**: $\mathcal{L}(T_M, \{t^*_j\}, \{t^*_{0,\text{on}i}\}, \{\delta^*_i\}) \neq \mathcal{L}(\lambda T_M, \{\lambda t^*_j\}, \{t^*_{0,\text{on}i}\}, \{\delta^*_i\})$ □

>>>>

**Activation Onset Times**: $t^*_{on,i} \sim \text{Normal}(\mu_{on}, \sigma_{on}^2)$

- **Dimensionless interpretation**: Gene $i$ activates at $t^*_{on,i}$ characteristic timescales after process initiation
- **Real-time interpretation**: Gene $i$ activates at $t^*_{on,i}/\alpha_{\text{ref}}$ hours after process start
- **Relative interpretation**: Gene $i$ activates at $(t^*_{on,i}/T_M) \times 100\%$ through the process

**Example interpretations:**
- $t^*_{on,i} = 1.5$, $T_M = 5.0$: Gene activates 30% through the process
- $t^*_{on,i} = -0.5$: Gene was already active 0.5 timescales before observation began

@@

- Posterior: $T_M \approx 6.0 \pm 0.5$
- Inferred characteristic timescale: $48/6 = 8$ hours
- Gene with $t^*_{on,i} = 2.0$: Activates at $2.0 \times 8 = 16$ hours
- Gene with $\delta^*_i = 1.5$: Active for $1.5 \times 8 = 12$ hours

@@

- $t^*_{on,i}$: When a gene's regulatory program activates (intrinsic timing)
- $\delta^*_i$: How long a gene's regulatory program lasts (intrinsic duration)

@@

\begin{align}
t^*_{on,i} &\sim \text{Normal}(1.5, 0.8^2) \quad \text{(fixed biological timing)} \\
\delta^*_i &\sim \text{LogNormal}(0.0, 0.45^2) \quad \text{(fixed biological duration)}
\end{align}

**NOT dependent on $T_M$:**
\begin{align}
t^*_{on,i} &\not\sim \text{Normal}(0.3 \times T_M, 0.8^2) \quad \text{(would create artificial scaling)}
\end{align}

@@

1. **Gene parameters are sampled independently**: $t^*_{on,i} \not\propto T_M$ and $\delta^*_i \not\propto T_M$

2. **Scaling cell times but not gene parameters changes piecewise phase assignments**:
   - Original: Cell $j$ in phase determined by comparing $t^*_j$ vs. $t^*_{on,i}$
   - Scaled: Cell $j$ in phase determined by comparing $\lambda t^*_j$ vs. $t^*_{on,i}$ (unchanged)

3. **The piecewise analytical solutions depend on comparisons like $t^*_j$ vs. $t^*_{on,i}$**:
   - Phase boundaries change when cell times are scaled but gene times are not
   - Different phase assignments lead to different analytical solution forms

4. **Therefore**: $\mathcal{L}(T_M, \{t^*_j\}, \{t^*_{on,i}\}, \{\delta^*_i\}) \neq \mathcal{L}(\lambda T_M, \{\lambda t^*_j\}, \{t^*_{on,i}\}, \{\delta^*_i\})$ □
